Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-04-30 manuscript/manuscript.tex
ms updates
ms uppppdaaates

@@

Due to the high computational burden of forward simulations over this wide
breadth of parameters, we modeled a single 50 megabase region in a population
of $N = 1000$ diploid individuals with a neutral variant mutation rate of
$10^{-8}$ and a recombination rate of $10^{-8}$ per basepair. This is roughly
analogous to a XXX chromosome of \emph{Drosophila melanogaster}; however with
this small population size and mutation rate, the population mutation rate
$\theta$ for the entire region leads to far fewer neutral sites to calculate
covariances and other statistics on than expected for a region this length in
\emph{D.  melanogaster}. Since our main goal is to understand how the dynamics
of statistics used in the paper and how they are affected by background
selection, fitness model, and trait architecture, we use population frequencies
rather than sampling frequencies.

@@

\subsubsection{The Effects of the Genetic Architecture}
\label{supp:genetic-arch}

@@

  \caption{The temporal covariances $\cov(\Delta p_5, \Delta p_t)$ from the
    onset of selection (generation 5) to a later time point $t$, which varies
    along the x-axis, across a variety of different trait additive genetic
    variances ($V_A$, columns) and number of sites contributing to the trait
    ($L$, rows). Each point is the temporal covariance averaged over 50
    replicate simulations; dark gray points are temporal covariances after the
    onset of selection, and light gray points are before. The red line is a
    loess smoothed curve through the covariances after the onset of selection.
    Selection on the trait was imposed through an exponential fitness function.}

@@

In our previous work, we did not investigate the affect of trait architecture
on our measure $G(t)$. Using the exponential fitness function simulations, we
also calculated $G(t)$ for each of the replicate simulations. We find that
across replicates, the $G(t)$ trajectories can vary considerably depending on
the number of sites ($L$) determining the trait's value (Supplementary Figure
\ref{suppfig:sim-expfit-Gs}). When a trait is monogenic ($L \approx 1$), $G(t)$
trajectories vary considerably across replicate lines, as certain lines may not
contain the few selected alleles (top row of Supplementary Figure
\ref{suppfig:sim-expfit-Gs}). However, with a polygenic trait, ($L \ge 100$),
the $G(t)$ trajectories across replicates are all quite close as each replicate
contains an abundance of alleles that affect the trait's value (bottom rows of
Supplementary Figure \ref{suppfig:sim-expfit-Gs}). Comparing the simulated
$G(t)$ replicate trajectories of Supplementary Figure
\ref{suppfig:sim-expfit-Gs} with the \textcite{Barghi2019-qy} $G(t)$
trajectories in Figure \ref{fig:figure-1}B, we again confirm a finding of
\textcite{Barghi2019-qy}: that there is considerable genetic redundancy among
beneficial alleles. We should note that our simplified simulation routines are
slightly different from the \textcite{Barghi2019-qy} study in that the burnin
populations are all independent; however we expect the same qualitative result.

@@

Additionally, we wanted to ensure that our temporal covariances and $G(t)$
trajectories were robust to more complicated, but realistic fitness models. To
this end, we simulated Gaussian stabilizing selection on a trait during burnin,
with two optima shift routines: (1) sudden optima shifts of $\mu_\text{sudden}
= \{0.1\sigma, 0.5\sigma, 1\sigma\}$, and (2) very graduate optima shifts of
$\mu_\text{gradual} = \{0.001\sigma, 0.01\sigma\}$ per generation using the
same two population simulation scheme described above. We used a polygenic
architecture for these simulations ($L = 1000$ trait affecting loci), and the
optima shift began at five generations after a $10N$ generation.

@@

  \label{gss-covs}
\end{figure}

@@

  \label{gss-covs}
\end{figure}

@@

  \label{gss-G}
\end{figure}

\subsubsection{Genetic Architecture and Convergence Correlations}

Using the same exponential fitness function simulations described above, we
also investigated how the convergence correlation is impacted by (1) genetic
architecture, (2) the design of the selection experiment, e.g. how many
individuals are selected for each line from the founding population, and (3)
the direction of selection across the two populations ``lines". After burning
in $N = 1000$ diploid populations for $10N$ generations, we simulated two
equally-sized lines of sizes $n = \{50, 500, 1,000\}$ diploids, and imposed
three selection schemes across different simulation runs. First, we imposed a
convergent selection scheme, where the populations undergo exponential
directional selection in the same direction. We expect that the convergent
correlation under this convergent scheme should be positive, as the two lines
should share some haplotypes carrying beneficial alleles, and these are
selected in the same direction across the two lines. Second, we imposed
divergent selection, where the two lines again undergo exponential directional
selection, except in different directions. Here, we expect the convergence
correlation to be negative, as haplotypes that increase the selected trait in
one population are beneficial in the upward selected line, but deleterious in
the downward selected line. Third, we have a control selection scheme, where
one line is selected and the other is not; this is akin to the control line in
the \textcite{Castro2019-uk} study (see Figure \ref{fig:figure-2}C). In this
case, we expect to see no convergence correlation, as only one line is being
selected. Finally, across these two-line simulation studies, we expect that
smaller sampled line sizes should show weaker convergent correlations, as the
probability that the same haplotypes are selected between the two lines
decreases with smaller size.

@@

Overall, our simulations confirm our hypotheses; see Supplementary Material
Figure \ref{suppfig:convergence-corrs}. We also find that in simulations where
we target a monogenic genetic architecture (i.e. the target number of
trait-affecting loci is $L=1$), the convergence correlations are generally much
weaker than those under a polygenic architecture. However, this effect is
mediated by the line population size; the difference in convergence correlation
between $L = 1$ and $L = 1000$ are more dissimilar when the line population
sizes are larger (compare the first column, last two rows). Like the
convergence correlations calculated on the \textcite{Barghi2019-qy} data, we
find in simulations convergence correlations decay through time. Overall, we
find that the convergence correlation is affected by both genetic architecture
and the size of the selected population lines.

We also wanted to test whether our we see similar convergence correlations
under Gaussian stabilizing selection. XXX We show these convergence
correlations for the sudden optima shifts in Supplementary Figure
\ref{suppfig:convergence-corrs-gss}. We found that for slow moving optima
shifts, the convergence correlations are generally too weak to be distinguished
from noise reliably, so these results are excluded (but see Supplementary
Materials Section XXX).

@@

In our analysis of the \textcite{Barghi2019-qy} data, we describe our statistic
$G(t)$ as an lower bound for two reasons: (1) the population is sequenced every
ten generations, meaning the temporal covariances between adjacent generations
cannot contribute to the numerator of $G(t)$ but contribute to the denominator,
and (2) the estimate of $G(t)$ ignores linked selection's contribution to the
variance in allele frequency change (this is the difference between $G(t)$ and
$G'(t)$, the latter of which includes these variance terms \cite{Buffalo2019-io}).
To verify that $G(t)$ estimated every ten generations is indeed a lower bound,
we used a simulation procedure similar to the exponential fitness function
simulations (described in Supplementary Material Section
\ref{supp:genetic-arch}), and calculated the temporal covariances and $G(t)$
both each generation, and every ten generations. Unlike the simulations
described in \ref{supp:genetic-arch}, we began selection at $10N$ generations,
and used trait $V_A = 0.01$ and $L = 1000$ sites affecting the trait.

First, comparing $G(t)$ when sampling population frequencies every generation
versus every ten generations, we confirm that the ten-generation block $G(t)$
is a lower bound of the true every generation $G(t)$ (red and blue lines in
Supplementary Figure \ref{suppfig:supp-blocks}A). Furthermore, since we control
the population size in our simulations at $N = 1000$ diploids, we know the
drift effective population size in the absence of selection. This allows us to
estimate $G(t)'$, which is a measure of $G(t)$ that accounts for the linked
selection's inflation of the variance in allele frequency change between two
generations (equation 26, \cite{Buffalo2019-io}). Plugging in the drift
effective population size $N_e = 1000$ into the expression for $G'(t)$ and
using the $\var(p_t - p_0)$ calculated for different $t$'s, we see that the
every generation $G(t)$ that does not account for linked selection's inflation
of $\var(\Delta p_t)$ does underestimate the true impact of linked selection as
expected (dashed gray line in Supplementary Figure \ref{suppfig:supp-blocks}A).

@@

\subsubsection{Background Selection's Effect on Temporal Covariance}

>>>>

\begin{figure}[!htb]
  \centering
  \includegraphics[width=\textwidth]{figures/figure-4-edited.pdf}

  \caption{Forward-in-time simulations of demonstrate how temporal covariance,
    $G(t)$ trajectories, and convergence correlations arise during optima
    shifts of two different magnitudes, under Gaussian stabilizing selection.
    (A) Trait means across 30 replicate before and after optima shifts (solid
    lines), for two different magnitudes (indicated by color). The optimal
    trait values are indicated by the purple and yellow dashed lines.  (B) Mean
    temporal covariance $\cov(\Delta p_5, \Delta p_t)$ across 30 simulation
    replicates, where $t$ varies along the x-axis (points), with a
    loess-smoothed average (solid line). (C) $G(t)$ trajectories through time,
    for 30 replicate simulations across two optima shifts. The solid line is a
    loess-smoothed average. (D) The convergence correlations between two
    populations split from a common population, that underwent either an optima
    shift in the same direction (converge) and opposite directions (diverge)
    at generation five. In subfigures (B), (C), and (D), directional selection
    begins at generation five, when the optima shifts; this is indicated by the
    vertical dashed red line.}

  \label{fig:figure-4}
\end{figure}

@@

Due to the high computational burden of forward simulations over this wide
breadth of parameters, we modeled a single 50 megabase region in a population
of $N = 1000$ diploid individuals with a neutral variant mutation rate of
$10^{-8}$ and a recombination rate of $10^{-8}$ per basepair. This is roughly
analogous to a quarter of an autosome of \emph{Drosophila melanogaster};
however with this small population size and mutation rate, the population
mutation rate $\theta$ for the entire region leads to far fewer neutral sites
to calculate covariances and other statistics on than expected for a region
this length in \emph{D.  melanogaster}. Since our main goal is to understand
the dynamics of statistics used in the paper and how they are affected by
different quantitative genetic fitness models, background selection, and trait
architecture, we use population frequencies rather than sampling frequencies.

@@

\subsubsection{The Effects of the Genetic Architecture under Exponential Directional Selection}
\label{supp:genetic-arch}

@@

  \caption{The temporal covariances $\cov(\Delta p_5, \Delta p_t)$ from the
    onset of selection (generation 5) to a later time point $t$, which varies
    along the x-axis, across a variety of different trait additive genetic
    variances ($V_A$, columns) and number of sites contributing to the trait
    ($L$, rows). Each point is the temporal covariance averaged over 50
    replicate simulations; dark gray points are temporal covariances after the
    onset of selection, and light gray points are before. The red line is a
    loess-smoothed curve through the covariances after the onset of selection.
    Selection on the trait was imposed through an exponential fitness function.}

@@

In our previous work, we did not investigate the affect of trait architecture
on our measure $G(t)$. Using the exponential fitness function simulations, we
also calculated $G(t)$ for each of the replicate simulations. We find that the
$G(t)$ trajectories can vary considerably across replicates depending on the
number of sites ($L$) determining the trait's value (Supplementary Figure
\ref{suppfig:sim-expfit-Gs}). When a trait is monogenic ($L \approx 1$), $G(t)$
trajectories vary considerably across replicate lines, as certain lines may not
contain the few selected alleles (top row of Supplementary Figure
\ref{suppfig:sim-expfit-Gs}). However, with a polygenic trait, ($L \ge 100$),
the $G(t)$ trajectories across replicates are all quite close as each replicate
contains an abundance of alleles that affect the trait's value (bottom rows of
Supplementary Figure \ref{suppfig:sim-expfit-Gs}). Comparing the simulated
$G(t)$ replicate trajectories of Supplementary Figure
\ref{suppfig:sim-expfit-Gs} with the \textcite{Barghi2019-qy} $G(t)$
trajectories in Figure \ref{fig:figure-1}B, we again confirm a finding of
\textcite{Barghi2019-qy}: that there is considerable genetic redundancy among
beneficial alleles. We should note that our simplified simulation routines are
slightly different from the \textcite{Barghi2019-qy} study in that the burnin
populations are all independent; however we expect the same qualitative result.

@@

@@

  \label{suppfig:gss-zbar}
\end{figure}

Additionally, we wanted to ensure that our temporal covariances and $G(t)$
trajectories were robust to more complicated, but realistic fitness models. To
this end, we also simulated Gaussian stabilizing selection (GSS) on a trait
during burnin, with two optima shift routines: (1) sudden optima shifts of
$\mu_\text{sudden} = \{0.1\sigma, 0.5\sigma, 1\sigma\}$, and (2) very graduate
optima shifts of $\mu_\text{gradual} = \{0.001\sigma, 0.01\sigma\}$ per
generation using the same two population simulation scheme described above. We
used a polygenic architecture for these simulations ($L = 1000$ trait affecting
loci), and the optima shift began at five generations after a $10N$ generation.
Across our GSS simulations, we see the expected selection response
(Supplementary Material Figure \ref{suppfig:gss-zbar}).

@@

  \caption{Mean temporal covariance ($\cov \Delta p_5, p_t$, with $t$ varying
    across the x-axis) across 30 replicate simulations (light gray points are
    before the onset of selection; dark gray points are after selection
    begins), under different Gaussian stabilizing selection with optima shift
  regimes. The solid red line is a loess-smoothed average of these points.}

  \label{suppfig:gss-covs}
\end{figure}

Overall, we see the same qualitative results under Gaussian stabilizing
selection with optima shifts as under exponential directional selection.
Stronger directional selection, here determined by larger sudden optima shifts
or larger gradual shifts per generation, lead to stronger temporal covariances
(Supplementary Materials Figure \ref{suppfig:covs}). Furthermore, we see a
stronger effect of linked selection, as measured by $G(t)$, under stronger
directional selection (Supplementary Material Figure \ref{suppfig:gss-G}).

@@

  \caption{$G(t)$ trajectories across 30 replicate Gaussian stabilizing
   selection with optima shift regimes. The solid red line is a loess-smoothed
   average across replicates.}

  \label{suppfig:gss-G}
\end{figure}

\subsubsection{Convergence Correlations}

Using the same exponential fitness function simulations described above, we
also investigated how the convergence correlation is impacted by (1) genetic
architecture, (2) the design of the selection experiment, e.g. how many
individuals are selected for each line from the founding population, and (3)
the direction of selection across the two populations ``lines". After burning
in $N = 1000$ diploid populations for $10N$ generations, we simulated two
equally-sized lines of sizes $n = \{50, 500, 1,000\}$ diploids, and imposed
three selection schemes across different simulation runs. First, we imposed a
convergent selection scheme, where the populations undergo exponential
directional selection in the same direction. We expect that the convergent
correlation under this convergent scheme should be positive, as the two lines
should share some haplotypes carrying beneficial alleles, and these are
selected in the same direction across the two lines. Second, we imposed
divergent selection, where the two lines again undergo exponential directional
selection, except in different directions. Here, we expect the convergence
correlation to be negative, as haplotypes that increase the selected trait in
one population are beneficial in the upward selected line, but deleterious in
the downward selected line. Third, we have a control selection scheme, where
one line is selected and the other is not; this is akin to the control line in
the \textcite{Castro2019-uk} study (see Figure \ref{fig:figure-2}C). In this
case, we expect to see no convergence correlation, as only one line is being
selected. Finally, across these two-line simulation studies, we expect that
smaller selection line sizes should show weaker convergent correlations, as the
probability that the same haplotypes are selected between the two lines
decreases with size.

@@

Overall, our simulations confirm our hypotheses; see Supplementary Material
Figure \ref{suppfig:convergence-corrs}. We also find that in simulations where
we target a monogenic genetic architecture (i.e. the target number of
trait-affecting loci is $L=1$), the convergence correlations are generally much
weaker than those under a polygenic architecture. However, this effect is
mediated by the line population size; the difference in convergence correlation
between $L = 1$ and $L = 1000$ are more dissimilar when the line population
sizes are larger (compare the first column, last two rows). Like the
convergence correlations calculated on the \textcite{Barghi2019-qy} data, we
find in simulations convergence correlations decay through time. Additionally,
populations selected in opposite directions lead to negative convergence
correlations, as expected. Overall, we find that the convergence correlation is
affected by both genetic architecture and the size of the selected population
lines.

We also wanted to test whether we see similar convergence correlations under
Gaussian stabilizing selection. In these simulations, rather than targeting a
particular $V_A$, we fix the trait mutation rate at $10^{-8}$ (thus region-wide
$\theta = 2000$). Like the exponential directional selection simulations, we
impose directional selection in the same direction across the two populations
(converge), different directions (diverge), and only in one population
(single). We also vary the type (gradual versus sudden) and magnitude of optima
shifts in the two populations. Overall, simulations show convergence
correlations for the sudden optima shifts in Supplementary Figure
\ref{suppfig:convergence-corrs-gss}. Importantly, optima shifts in the opposite
direction cause negative convergence correlations. We found that for slow
moving optima shifts, the convergence correlations are generally too weak to be
distinguished from zero reliably, (top row of Supplementary Material Figure
\ref{suppfig:convergence-corrs-gss}).

@@

In our analysis of the \textcite{Barghi2019-qy} data, we describe our statistic
$G(t)$ as a lower bound for two reasons: (1) the population is sequenced every
ten generations, meaning the temporal covariances between adjacent generations
cannot contribute to the numerator of $G(t)$ but contribute to the denominator,
and (2) the estimate of $G(t)$ ignores linked selection's contribution to the
variance in allele frequency change (this is the difference between $G(t)$ and
$G'(t)$, the latter of which includes these variance terms
\parencite{Buffalo2019-io}).  To verify that $G(t)$ estimated every ten
generations is indeed a lower bound, we used a simulation procedure similar to
the exponential fitness function simulations (described in Supplementary
Material Section \ref{supp:genetic-arch}), and calculated the temporal
covariances and $G(t)$ both each generation, and every ten generations. Unlike
the simulations described in \ref{supp:genetic-arch}, we began selection at
$10N$ generations, and used trait $V_A = 0.01$ and targeted $L = 1000$ sites
affecting the trait.

First, comparing $G(t)$ when sampling population frequencies every generation
versus every ten generations, we confirm that the ten-generation block $G(t)$
is a lower bound of the $G(t)$ trajectory when sampling is every generation
(red and blue lines in Supplementary Figure \ref{suppfig:supp-blocks}A).
Furthermore, since we control the population size in our simulations at $N =
1000$ diploids, we know the drift effective population size in the absence of
selection. This allows us to estimate $G(t)'$, which is a measure of $G(t)$
that accounts for the linked selection's inflation of the variance in allele
frequency change between two generations (equation 26, \cite{Buffalo2019-io}).
Plugging in the drift effective population size $N_e = 1000$ into the
expression for $G'(t)$ and using the $\var(p_t - p_0)$ calculated for different
$t$'s, we see that the every generation $G(t)$ that does not account for linked
selection's inflation of $\var(\Delta p_t)$ does underestimate the true impact
of linked selection as expected (dashed gray line in Supplementary Figure
\ref{suppfig:supp-blocks}A).

@@

\subsubsection{Background Selection}

In our previous work, \textcite{Buffalo2019-io}, we did not investigate whether
background selection can lead to temporal autocovariance. Here, using
Forward-in-time simulations, we find that background selection can indeed
generate temporal autocovariance and lead to convergence correlations when
deleterious haplotypes are shared between populations and both removed by
selection.

We simulated background selection in a 50 megabase region, where deleterious
alleles are randomly introduced by mutation. Following background selection
literature
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}, we
parameterize the mutation rate as the total number of deleterious mutations
introduced per diploid genome, per generation, and simulate values $U = \{0.5,
1.0, 1.5\}$. We also vary the strength of selection against the deleterious
mutations, $s = \{0.0, 0.01, 0.05, 0.1\}$ (where $s=0$ is a neutral control),
as well as different recombination rates ($r_{bp} = \{10^{-7}, 10^{-8}\}$).
Like other simulations, we burnin the population for $10N$ generations under
backgrounds selection. Overall, we find background selection does create
temporal covariance (Supplementary Material Figure
\ref{suppfig:supp-bgs-covs}), which are stronger under (1) higher deleterious
mutation rates and (2) larger selection coefficients. This latter point
initially seems at odds with background selection theory, as the level of
pairwise diversity in a region under background selection is invariant with
respect to the selection coefficient. However, looking at the background
selection $G(t)$ trajectories, we find that over time, the impact of linked
selection created by background selection appears to trend towards an
equilibrium in the $r_{BP} = 10^{-7}$ subfigure, and reaches an equilibrium in
the $r_{BP} = 10^{-8}$ subfigure (Supplementary Material Figure
\ref{suppfig:supp-bgs-g}). However, further future work is needed to fully
explore and understand the temporal covariance dynamics of background
selection.

Note about fixations

  \caption{The temporal covariances, $\cov(\Delta p_{50}, \Delta p_t)$ (where $t$
    varies along the x-axis) created by background selection. Unlike
    directional selection figures, where we choose the reference generation to
    be the first generation after the onset of selection, here we choose an
    arbitrary reference generation (generation 50). The symmetry of temporal
    covariance around the reference generation, is expected, since unlike
    directional selection the level of additive genetic variance for fitness has
    hit mutation-selection-drift balance.}
